Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-30 paper_nips/paper.tex
removed heavy subscripts from biases by naming some of them c

We were further interested in the utility of the optimal low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classifcation problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
81 right-handed healthy participants
(3 not included in present analyses due to incomplete data)
without psychiatric or
neurological history participated in four fMRI sessions acquired under
different experimental paradigms.
The functional maps were warped into
the MNI space and resampled to isotropic 3mm resolution.
Whole-brain EPI data were acquired with the same Siemens Trio with a 32
channel head coil (TR=2400ms, TE=30ms, flip angle=60°, in-plane
FOV=$192\textrm{mm}\times192\textrm{mm}$, 40 slices, 3.0mm isotropic voxels).
Standard preprocessing was performed with Nipype \cite{gorgo11}, including
slice timing, motion correction, alignment, and spatial normalization.
activity maps were spatially smoothed by
a Gaussian kernel of 5mm (FWHM).
Analogous to HCP data, the second task dataset incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

@@

\begin{eqnarray}
  \begin{split}
    \mathbf{z} &= \mathbf{W_0} \mathbf{x} + \mathbf{b_{W_0}} \\
    \mathbf{x'} &= \mathbf{W_1} \mathbf{z} + \mathbf{b_{W_1}}
  \end{split}
  \label{eq:autoenc}
\end{eqnarray}

where $\mathbf{x \in \mathbb{R}^{d}}$ denotes the vector of $d=79,941$
voxel values from each
rest map,
$\mathbf{z \in \mathbb{R}^{n}}$ is the $n$ hidden dimensions (i.e.,
distributed neural activity patterns), and 
$\mathbf{x' \in \mathbb{R}^{d}}$
is the reconstruction vector of the original activity map
from the hidden variables. 
% XXX use bold letters for vectors and matrices
Further, $\mathbf{W_0}$ denotes the weight matrix that
transforms
from input space into the hidden space (encoder),
$\mathbf{W_1}$ is the weight matrix for back-projection
from the hidden variables to the
output space (decoder).
$\mathbf{b_{W_0}}$ and $\mathbf{b_{W_1}}$ are bias vectors.
Note that $\mathbf{W_0}$ and $\mathbf{W_1}$ are tied,
$\mathbf{W_0} = \mathbf{W_1^T}$.
The optimal model parameters $\mathbf{W_0, b_{W_0}, b_{W_1}}$ are found by
minimizing the imperfection of reconstruction according to squared error

@@

\begin{eqnarray}
  \begin{split}
    P(Y=i|\mathbf{x, V_0,V_1,b_{V_0}, b_{V_1}}) &= softmax_i(\mathbf{V_1 (V_0 x + b_{V_0}) + b_{V_1}}) \\
    &= \frac {e^{\mathbf{V_{1i} (V_{0i} x + b_{V_{0i}}) + b_{V_{1i}}}}} {\sum_j e^{\mathbf{V_{1j} (V_{0j} x + b_{V_{0j}}) + b_{V_{1j}}}}}            
  \end{split}
  \label{eq:lr}
\end{eqnarray}

where the matrix $\mathbf{V_0 \in \mathbb{R}^{dxn}}$
transforms the input $\mathbf{x \in \mathbb{R}^{d}}$
into $n$ latent components
and the matrix $\mathbf{V_1 \in \mathbb{R}^{nxc}}$
projects the latent components
onto hyperplanes that reflect $c$ class probabilities.
$\mathbf{b_{V_0}}$ and $\mathbf{b_{V_1}}$ are corresponding
bias vectors.
The loss function is given by

\begin{eqnarray}
  \begin{split}
    {\mathcal{L_{LR}}}(\mathbf{x, y}) = \frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|\mathbf{x^{(i)}; V_1 (V_0 x + b_{V_0}) + b_{V_1}})
\end{split}
\label{eq:lr_loss}
\end{eqnarray}

@@

In so doing, we search for the combined model parameters
$\theta=\{\mathbf{V_0,V_1,b_{V_0}, b_{V_1}, b_{W_0}, b_{W_1}}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task classification.
${\mathcal{L_{AE}}}$ is devided by ${N_{X_{rest}}}$ to equilibrate both
loss terms to the same order of magnitude.
For parameter shrinkage, we
impose a combination of $\ell_1$ and $\ell_2$ penalties
$\forall p \in \theta$ (i.e., elasticnet).

>>>>

We were further interested in the utility of the optimal low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classifcation problem of another large sample.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
81 right-handed healthy participants
(3 not included in present analyses due to incomplete data)
without psychiatric or
neurological history participated in four fMRI sessions acquired under
different experimental paradigms.
The functional maps were warped into
the MNI space and resampled to isotropic 3mm resolution.
Whole-brain EPI data were acquired with the same Siemens Trio with a 32
channel head coil (TR=2400ms, TE=30ms, flip angle=60°, in-plane
FOV=$192\textrm{mm}\times192\textrm{mm}$, 40 slices, 3.0mm isotropic voxels).
Standard preprocessing was performed with Nipype \cite{gorgo11}, including
slice timing, motion correction, alignment, and spatial normalization.
Activity maps were spatially smoothed by
a Gaussian kernel of 5mm (FWHM).
Analogous to HCP data, the second task dataset incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

@@

\begin{eqnarray}
  \begin{split}
    \mathbf{z} &= \mathbf{W_0} \mathbf{x} + \mathbf{b_0} \\
    \mathbf{x'} &= \mathbf{W_1} \mathbf{z} + \mathbf{b_1}
  \end{split}
  \label{eq:autoenc}
\end{eqnarray}

where $\mathbf{x \in \mathbb{R}^{d}}$ denotes the vector of $d=79{,}941$
voxel values from each
rest map,
$\mathbf{z \in \mathbb{R}^{n}}$ is the $n$\suggestremove{ hidden dimensions 
(i.e.,
distributed neural activity patterns)}
\suggestadd{-dimensional hidden state}, and 
$\mathbf{x' \in \mathbb{R}^{d}}$
is the reconstruction vector of the original activity map
from the hidden variables. 
% XXX use bold letters for vectors and matrices
Further, $\mathbf{W_0}$ denotes the weight matrix that
transforms
from input space into the hidden space (encoder),
$\mathbf{W_1}$ is the weight matrix for back-projection
from the hidden variables to the
output space (decoder).
$\mathbf{b_0}$ and $\mathbf{b_1}$ are bias vectors.
Note that $\mathbf{W_0}$ and $\mathbf{W_1}$ are tied,
$\mathbf{W_0} = \mathbf{W_1^T}$.
The optimal model parameters $\mathbf{W_0, b_0, b_1}$ are found by
minimizing the imperfection of reconstruction according to squared error

@@

\begin{eqnarray}
  \begin{split}
    P(Y=i|\mathbf{x, V_0,V_1,c_0, c_1}) &= softmax_i(\mathbf{V_1 (V_0 x + c_0) + c_1}) \\
    &= \frac {e^{\mathbf{V_{1i} (V_{0i} x + b_{V_{0i}}) + b_{V_{1i}}}}} {\sum_j e^{\mathbf{V_{1j} (V_{0j} x + b_{V_{0j}}) + b_{V_{1j}}}}}            
  \end{split}
  \label{eq:lr}
\end{eqnarray}

where the matrix $\mathbf{V_0 \in \mathbb{R}^{dxn}}$
transforms the input $\mathbf{x \in \mathbb{R}^{d}}$
into $n$ latent components
and the matrix $\mathbf{V_1 \in \mathbb{R}^{nxc}}$
projects the latent components
onto hyperplanes that reflect $c$ class probabilities.
$\mathbf{c_0}$ and $\mathbf{c_1}$ are corresponding
bias vectors.
The loss function is given by

\begin{eqnarray}
  \begin{split}
    {\mathcal{L_{LR}}}(\mathbf{x, y}) = \frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|\mathbf{x^{(i)}; V_1 (V_0 x + c_0) + c_1})
\end{split}
\label{eq:lr_loss}
\end{eqnarray}

@@

In so doing, we search for the combined model parameters
$\theta=\{\mathbf{V_0,V_1,c_0, c_1, b_0, b_1}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task classification.
${\mathcal{L_{AE}}}$ is devided by ${N_{X_{rest}}}$ to equilibrate both
loss terms to the same order of magnitude.
For parameter shrinkage, we
impose a combination of $\ell_1$ and $\ell_2$ penalties
$\forall p \in \theta$ (i.e., elasticnet).
